---
title: "Reproducible Research Project"
author: "Hewan"
date: "11/23/2020"
output: html_document
---
#

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

#cleaning the R consol and creating the directory

```{r}
ls()
remove(list=ls())
graphics.off()
getwd()
setwd("C:/Users/Inspiron 5537pro/Desktop/Project/Reproducible_Research")

```

#Data and tye for this project
*The data is  downloaded from the course web site:

*Dataset: Activity Monitoring Data [52K]

*The variables included in this dataset are:

*steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

*date: The date on which the measurement was taken in YYYY-MM-DD format

*interval: Identifier for the 5-minute interval in which measurement was taken

*The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
#Loading and preprocessing the data

*Loading the Data

```{r, echo=TRUE}
library(ggplot2)
library(plyr)
activity <- read.csv("activity.csv")
```
Processing the Data
```{r}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
```
##pulling data without nas
```{r}
clean <- activity[!is.na(activity$steps),]
```
#mean total number of steps taken per day
## summarizing total steps per date
```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
```
#Make a histogram of the total number of steps taken each day

```{r}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```
Calculate and report the mean and median of the total number of steps taken per day
```{r}
#mean
as.integer(mean(sumTable$Steps))
#median
```
The result indicates that The average number of steps taken each day was 10766 steps.

and

The median number of steps taken each day was 10765 steps.

#What is the average daily activity pattern?

Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
library(plyr)
library(ggplot2)
##pulling data without nas
clean <- activity[!is.na(activity$steps),]

##create average number of steps per interval
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))

##Create line plot of average number of steps per interval
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ ggtitle("Average Number of Steps per Interval")
```
The following steps is to identify across all the days in the dataset, contains the maximum number of steps
```{r}
##Maximum steps by interval
maxSteps <- max(intervalTable$Avg)
##Which interval contains the maximum average number of steps
intervalTable[intervalTable$Avg==maxSteps,1]
```
The maximum number of steps for a 5-minute interval was 206 steps.

The 5-minute interval which had the maximum number of steps was the 835 interval



Calculate and report the total number of missing values in the dataset 

#Imputing missing values 
```{r}
##Number of NAs in original data set
nrow(activity[is.na(activity$steps),])
```
The total number of rows with steps is 2304.

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

My strategy for filling in NAs will be to substitute the missing steps with the average 5-minute interval based on the day of the week.
```{r}
## Create the average number of steps per weekday and interval
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))

## Create dataset with all NAs for substitution
nadata<- activity[is.na(activity$steps),]
## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.

## Reorder the new substituded data in the same format as clean data set
```{r}
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

##Merge the NA averages and non NA data together
mergeData <- rbind(clean, newdata2)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

##Create sum of steps per date to compare with step 1
```{r}
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")

## Mean of Steps with NA data taken care of
as.integer(mean(sumTable2$Steps))

## Median of Steps with NA data taken care of
as.integer(median(sumTable2$Steps))

## Creating the histogram of total steps per day, categorized by data set to show impact
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```
The new mean of the imputed data is 10821 steps compared to the old mean of 10766 steps. That creates a difference of 55 steps on average per day.

The new median of the imputed data is 11015 steps compared to the old median of 10765 steps. That creates a difference of 250 steps for the median.

However, the overall shape of the distribution has not changed.
Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

## Create new category based on the days of the week
```{r}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```

Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
library(lattice) 

## Warning: package 'lattice' was built under R version 3.1.3

## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),main="Average Steps per Interval Based on Type of Day", ylab="Average Number of Steps", xlab="Interval")
```
Yes, the step activity trends are different based on whether the day occurs on a weekend or not. This may be due to people having an increased opportunity for activity beyond normal work hours for those who work during the week.